### 2021-01-07 ch03_esmvaltool/esmvaltool.tex
Finished chapter 3

\begin{table}[t]
  \centering
  \begin{tabulary}{\columnwidth}{L L L}
    \toprule
    Type & Name & Description \\
    \midrule
    Preprocessor & \emph{amplitude} & Amplitude of periodic phenomena (\eg{}
    diurnal or seasonal cycles) \\
    \cmidrule(l){2-3}
    & \emph{land/sea fraction weighting} & Weighting of fields based on
    the land or sea fraction of the respective grid cells \\
    \cmidrule(l){2-3}
    & \emph{trend} & Linear trend (slope of ordinary least squares regression)
    \\
    \midrule
    Derived variable & \emph{asr} & Absorbed shortwave radiation \\
    \cmidrule(l){2-3}
    & \emph{co2s} & Atmospheric \ac{CO2} concentration at
    surface \\
    \cmidrule(l){2-3}
    & \emph{et} & Evapotranspiration \\
    \cmidrule(l){2-3}
    & \emph{rlntcs} & Clear-sky net \acf{TOA} longwave radiation \\
    \cmidrule(l){2-3}
    & \emph{rsntcs} & Clear-sky net \ac{TOA} shortwave radiation \\
    \cmidrule(l){2-3}
    & \emph{uajet} & Position of austral jet stream \\
    \bottomrule
  \end{tabulary}
  \caption{Summary of changes and features contributed to the
    \acf{ESMValCore}.}
  \label{tab:03:changes_to_esmvalcore}
\end{table}

@@

Apart from new preprocessor functions, six derivation scripts for the
non-\ac{CMOR} variables \emph{asr}, \emph{co2s}, \emph{et}, \emph{rlntcs},
\emph{rsntcs} and \emph{uajet} have been added. The absorbed shortwave
radiation (\emph{asr}) is defined as the difference in the incoming \ac{TOA}
shortwave radiation and the outgoing \ac{TOA} shortwave radiation. The
atmospheric \ac{CO2} concentration at surface (\emph{co2s}) can be calculated
from the pressure level-dependent atmospheric \ac{CO2} concentration and the
surface air pressure using interpolation. An example illustrating \emph{co2s}
is given in \cref{fig:03:co2s}. \Cref{fig:03:co2s:a} shows the \acs{CMIP}6
\ac{MMM} of the global \emph{co2s} for the months \acf{JJA} of the year 2014 in
the emission-driven historical simulation. As expected, the map shows high
atmospheric \ac{CO2} concentrations over large metropolitan areas with high
\ac{CO2} emissions (\eg{} over North America, the Arabian Peninsula and East
Asia). Moreover, regions with high photosynthetic activity (\eg{} the boreal
forests in the Northern hemisphere summer and the tropical rainforests in South
America) exhibit smaller \ac{CO2} concentrations. \Cref{fig:03:co2s:b} shows
the monthly-mean \emph{co2s} at Mauna Loa, Hawaii from 1958 to 2014 for
observations performed at the Mauna Loa Observatory \autocite{Keeling2005}
(thick black line) and emission-driven historical simulations from four
\acs{CMIP}6 models (remaining colored lines). Apart from the model CNRM-ESM2-1,
the simulated atmospheric \ac{CO2} concentrations show the expected seasonal
cycle present in the Keeling Curve (see \cref{fig:02:keeling_curve}). All
climate models correctly simulate the increase in \emph{co2s} over the years.
However, their are some differences in the absolute values with the model
BCC-CSM2-MR showing the largest deviations.

@@

\begin{figure}[t]
  \centering
  \begin{subfigure}[t]{\SubfigureWidth{}}
    \includegraphics[width=\columnwidth]{ch03_esmvaltool/figs/cox18_psi.pdf}
    \caption{}
    \label{fig:03:cox18:a}
  \end{subfigure}
  ~
  \begin{subfigure}[t]{\SubfigureWidth{}}
    \includegraphics[width=\columnwidth]{ch03_esmvaltool/figs/cox18_ec.pdf}
    \caption{}
    \label{fig:03:cox18:b}
  \end{subfigure}
  \caption{Emergent constraint on the \acf{ECS} from global temperature
    variability \autocite{Cox2018}. Magenta colors indicate high sensitivity
    climate models ($\lambda < 1 \unit{W m^{-2} K^{-1}}$) and green colors
    indicate low sensitivity climate models ($\lambda > 1 \unit{W m^{-2}
    K^{-1}}$). Both panels were created with the \acs{ESMValTool} and are
    similar to \textcite{Cox2018}. (a) Temperature variability metric $\Psi$
    versus time for historical simulations from \acs{CMIP}5 models (colored
    lines) and observations from the HadCRUT4 dataset (black circles)
    \autocite{Morice2012}. (b) Emergent relationship between \acs{ECS} and the
    temperature variability metric $\Psi$. The black dot-dashed line shows the
    linear regression across the climate model ensemble with its associated
    uncertainties indicated by the black dashed lines (standard prediction
    error; see \cref{eq:02:spe}). Blue lines show the observational constraint
    from HadCRUT4 with its best estimate (dot-dashed line) and standard
    deviation (dashed lines). The letters represent individual \acs{CMIP}5
    models (see \textcite{Cox2018} for details).}
  \label{fig:03:cox18}
\end{figure}

\begin{figure}[t]
  \centering
  \begin{subfigure}[t]{0.34\columnwidth}
    \includegraphics[width=\columnwidth]{ch03_esmvaltool/figs/flato13_942a.pdf}
    \caption{}
    \label{fig:03:flato13:a}
  \end{subfigure}
  ~
  \begin{subfigure}[t]{0.34\columnwidth}
    \includegraphics[width=\columnwidth]{ch03_esmvaltool/figs/flato13_942b.pdf}
    \caption{}
    \label{fig:03:flato13:b}
  \end{subfigure}
  ~
  \begin{subfigure}[t]{0.24\columnwidth}
    \raisebox{7.5mm}{\includegraphics[width=\columnwidth]{
      ch03_esmvaltool/figs/flato13_942_legend.pdf}}
  \end{subfigure}
  \caption{Relationships between climate metrics within the \acs{CMIP}5 climate
    model ensemble. Both panels were created with the \acs{ESMValTool} and are
    similar to \textcite{Flato2013}. (a) \Acf{GSAT} versus the \acf{ECS} for
    \acs{CMIP}5 models for the period \range{1961}{1990} (larger symbols) and
    for pre-industrial control runs (smaller symbols). (b) \acf{TCR} versus
    \acs{ECS} for \acs{CMIP}5 models. The black line shows a linear fit
    between \acs{TCR} and \acs{ECS}.}
\end{figure}

- total number of changed lines of code

- Added recipes and diagnostics for ESMValTool + table

* Anav13

* MDER

* TCR

* ECS

* Schlund20jgr

* Schlund20ESD

* Flato13

>>>>

\begin{table}[t]
  \centering
  \begin{tabulary}{\columnwidth}{L L L}
    \toprule
    Type & Name & Description \\
    \midrule
    Preprocessor & \emph{amplitude} & Amplitude of periodic phenomena (\eg{}
    diurnal or seasonal cycles) \\
    \cmidrule(l){2-3}
    & \emph{land/sea fraction weighting} & Weighting of fields based on
    the land or sea fraction of the respective grid cells \\
    \cmidrule(l){2-3}
    & \emph{trend} & Linear trend (slope of ordinary least squares regression)
    \\
    \midrule
    Derived variable & \emph{asr} & Absorbed shortwave radiation \\
    \cmidrule(l){2-3}
    & \emph{co2s} & Atmospheric \ac{CO2} concentration at
    surface \\
    \cmidrule(l){2-3}
    & \emph{et} & Evapotranspiration \\
    \cmidrule(l){2-3}
    & \emph{rlntcs} & Clear-sky net \acf{TOA} longwave radiation \\
    \cmidrule(l){2-3}
    & \emph{rsntcs} & Clear-sky net \ac{TOA} shortwave radiation \\
    \cmidrule(l){2-3}
    & \emph{uajet} & Position of austral jet stream \\
    \bottomrule
  \end{tabulary}
  \caption{Summary of new preprocessor functions and variable derivation
    scripts contributed to the \acf{ESMValCore} by the author of this thesis.}
  \label{tab:03:changes_to_esmvalcore}
\end{table}

@@

Apart from new preprocessor functions, six derivation scripts for the
non-\ac{CMOR} variables \emph{asr}, \emph{co2s}, \emph{et}, \emph{rlntcs},
\emph{rsntcs} and \emph{uajet} have been added. The absorbed shortwave
radiation (\emph{asr}) is defined as the difference in the incoming \ac{TOA}
shortwave radiation and the outgoing \ac{TOA} shortwave radiation. The
atmospheric \ac{CO2} concentration at surface (\emph{co2s}) can be calculated
from the pressure level-dependent atmospheric \ac{CO2} concentration and the
surface air pressure using interpolation. An example illustrating \emph{co2s}
is given in \cref{fig:03:co2s}. \Cref{fig:03:co2s:a} shows the \acs{CMIP}6
\ac{MMM} of the global \emph{co2s} for the months \acf{JJA} of the year 2014 in
the emission-driven historical simulation. As expected, the map shows high
atmospheric \ac{CO2} concentrations over large metropolitan areas with high
\ac{CO2} emissions (\eg{} over North America, the Arabian Peninsula and East
Asia). Moreover, regions with high photosynthetic activity (\eg{} the boreal
forests in the Northern hemisphere summer and the tropical rainforests in South
America) exhibit smaller \ac{CO2} concentrations. \Cref{fig:03:co2s:b} shows
the monthly-mean \emph{co2s} at Mauna Loa, Hawaii from 1958 to 2014 for
observations performed at the Mauna Loa Observatory \autocite{Keeling2005}
(thick black line) and emission-driven historical simulations from four
\acs{CMIP}6 models (remaining colored lines). Apart from the model CNRM-ESM2-1,
the simulated atmospheric \ac{CO2} concentrations show the expected seasonal
cycle present in the Keeling Curve (see \cref{fig:02:keeling_curve}). All
climate models correctly simulate the increase in \emph{co2s} over the years.
However, there are some differences in the absolute values with the model
BCC-CSM2-MR showing the largest deviations.

@@

In addition to the analyses presented in
\cref{ch:04:assessment_climate_metrics,ch:05:paper_ecs,ch:06:paper_gpp},
several more recipes and \ac{CMOR}ization scripts for observational data have
been added to the main \ac{ESMValTool} repository on GitHub. Moreover, multiple
minor changes and additions have been implemented to improve the code base
and/or to fix bugs. In total, the author of this study has added 88668 lines of
code and removed 47816 lines of code from the \ac{ESMValTool} repository (as of
December 2020). Parts of these implementations are already published in the
scientific documentation of the \ac{ESMValTool}, which covers large-scale
diagnostics \autocite{Eyring2020}, diagnostics for emergent constraints and
future projections \autocite{Lauer2020} and diagnostics for extreme events and
regional evaluation \autocite{Weigel2020}.

The implemented \ac{CMOR}ization scripts (also referred to as
\emph{\ac{CMOR}izers}) allow the processing of arbitrary observational data
with the \ac{ESMValTool}. For this, the raw observational datasets are
reformatted in such way that they adhere to the \ac{CMOR} standards. Since the
\ac{CMOR}ization scripts are publicly available like any other code in the
\ac{ESMValTool} and \ac{ESMValCore} repositories, the entire scientific
community can use them to \ac{CMOR}ize the observational data. A complete list
of all eleven newly implemented \ac{CMOR}izers including the main reference for
the corresponding observational datasets is shown in
\cref{tab:03:changes_to_esmvaltool}.

Due to an exhaustive overhaul of the \ac{ESMValTool} version 2, older recipes
from version 1 \autocite{Eyring2016a} cannot be used directly in the latest
release. Porting recipes from the old version to the current version requires
substantial changes on the recipes themselves, but also on the corresponding
diagnostic scripts. As of December 2020, the author of this thesis has ported
two recipes originally written by other authors to the new version of
\ac{ESMValTool}: \emph{recipe\_anav13jclim.yml} and
\emph{recipe\_wenzel16jclim.yml}. \emph{recipe\_anav13jclim.yml} reproduces the
analysis of \textcite{Anav2013}, who evaluate carbon cycle-related variables
for climate models of the \acs{CMIP}5 ensemble. This includes climatologies,
trends and variabilities of important quantities of the terrestrial and oceanic
carbon cycle like carbon fluxes, carbon reservoir sizes and vegetation
distributions. \emph{recipe\_wenzel16jclim.yml} includes the \ac{MDER} analysis
performed by \textcite{Wenzel2016a} to constrain future changes in the position
of the austral jet stream. Due to its flexible code, the corresponding
diagnostic scripts can be used to apply the \ac{MDER} method to arbitrary
target variables and predictors.

\begin{table}[t]
  \centering
  \begin{tabulary}{\columnwidth}{L L L}
    \toprule
    Type & Name & Main reference \\
    \midrule
    \ac{CMOR}izer & CRU & \autocite{Harris2014} \\
    & CT2019 & \autocite{Jacobson2020} \\
    & GCP & \autocite{Friedlingstein2019} \\
    & HWSD & \autocite{Wieder2014} \\
    & JMA-TRANSCOM & \autocite{Maki2010} \\
    & LAI3g & \autocite{Zhu2013} \\
    & LandFlux-EVAL & \autocite{Mueller2013} \\
    & MLS-Aura & \autocite{Read2015} \\
    & MTE & \autocite{Jung2011} \\
    & NDP & \autocite{Gibbs2006} \\
    & Scripps-\ac{CO2} & \autocite{Keeling2005} \\
    \midrule
    Recipe & \emph{bock20jgr} (figures 8 and 10) & \autocite{Bock2020} \\
    & \emph{cox18nature} & \autocite{Cox2018} \\
    & \emph{ecs} & \autocite{Gregory2004} \\
    & \emph{flato13ipcc} (figure 9.42) & \autocite{Flato2013} \\
    & \emph{schlund20esd} & \autocite{Schlund2020a} \\
    & \emph{schlund20jgr\_gpp\_abs\_rcp85} & \autocite{Schlund2020} \\
    & \emph{schlund20jgr\_gpp\_change\_1pct} & \autocite{Schlund2020} \\
    & \emph{schlund20jgr\_gpp\_change\_rcp85} & \autocite{Schlund2020} \\
    & \emph{tcr} & \autocite{Gregory2008} \\
    \bottomrule
  \end{tabulary}
  \caption{Summary of new \acs{CMOR}ization scripts for observational data
    and recipes contributed to the \acf{ESMValTool} by the author of this
    thesis. The corresponding file names of the recipes in the
    \acs{ESMValTool} repository are given by
    \emph{recipe\_(name\_in\_table).yml}.}
  \label{tab:03:changes_to_esmvaltool}
\end{table}

In addition to the ported recipes, several new recipes and diagnostic scripts
have been contributed to the public \ac{ESMValTool} repository (see
\cref{tab:03:changes_to_esmvaltool} for a summary). Some of these include the
analyses performed in other chapters of this thesis, namely
\emph{recipe\_bock20jgr.yml} (\cref{ch:04:assessment_climate_metrics}),
\emph{recipe\_schlund20esd.yml} (\cref{ch:05:paper_ecs}) and
\emph{recipe\_schlund20jgr\_*.yml} (\cref{ch:06:paper_gpp}).
\emph{recipe\_ecs.yml} and \emph{recipe\_tcr.yml} contain diagnostic scripts
that are able to calculate the climate metrics \ac{ECS} and \ac{TCR} for
arbitrary model output (see \cref{sec:02:climate_sensitivity}). Corresponding
plots are shown in \cref{fig:02:gregory_regression,fig:02:tcr}.
\emph{recipe\_cox18nature.yml} reproduces the analysis of \textcite{Cox2018},
who introduce an emergent constraint in \ac{ECS} based on a global temperature
variability metric $\Psi$. \Cref{fig:03:cox18:a} shows $\Psi$ over the past 85
years for \acs{CMIP}5 models and observations form HadCRUT4
\autocite{Morice2012}. Since high sensitivity climate models exhibit higher
values of $\Psi$ than low sensitivity models, an emergent relationship between
\ac{ECS} and $\Psi$ can be established (\cref{fig:03:cox18:b}). With
observations of $\Psi$, \ac{ECS} can eventually be constrained to $\left( 2.8
\pm 0.6 \right) \unit{K}$ ($66 \unit{\%}$ confidence range). More details on
this including the calculation of $\Psi$ are given in \cref{subsec:05:cox}.

\begin{figure}[t]
  \centering
  \begin{subfigure}[t]{\SubfigureWidth{}}
    \includegraphics[width=\columnwidth]{ch03_esmvaltool/figs/cox18_psi.pdf}
    \caption{}
    \label{fig:03:cox18:a}
  \end{subfigure}
  ~
  \begin{subfigure}[t]{\SubfigureWidth{}}
    \includegraphics[width=\columnwidth]{ch03_esmvaltool/figs/cox18_ec.pdf}
    \caption{}
    \label{fig:03:cox18:b}
  \end{subfigure}
  \caption{Emergent constraint on the \acf{ECS} from global temperature
    variability \autocite{Cox2018}. Magenta colors indicate high sensitivity
    climate models ($\lambda < 1 \unit{W m^{-2} K^{-1}}$) and green colors
    indicate low sensitivity climate models ($\lambda > 1 \unit{W m^{-2}
      K^{-1}}$). Both panels were created with the \acs{ESMValTool} using
    \emph{recipe\_cox18nature.yml} and are similar to \textcite{Cox2018}. (a)
    Temperature variability metric $\Psi$ versus time for historical
    simulations from \acs{CMIP}5 models (colored lines) and observations from
    the HadCRUT4 dataset (black circles) \autocite{Morice2012}. (b) Emergent
    relationship between \acs{ECS} and the temperature variability metric
    $\Psi$. The black dot-dashed line shows the linear regression across the
    climate model ensemble with its associated uncertainties indicated by the
    black dashed lines (standard prediction error; see \cref{eq:02:spe}). Blue
    lines show the observational constraint from HadCRUT4 with its best
    estimate (dot-dashed line) and standard deviation (dashed lines). The
    letters represent individual \acs{CMIP}5 models (see \textcite{Cox2018}
    for details).}
  \label{fig:03:cox18}
\end{figure}

An important class of recipes in the \ac{ESMValTool} covers international climate assessments. One such example is \emph{recipe\_flato13ipcc.yml}, which reproduces large parts of the climate model evaluation performed in chapter 9 of the \ac{IPCC}'s \ac{AR5} \autocite{Flato2013}. For this recipe, the two-panel figure 9.42 has been added, which shows the relationship between the historical and pre-industrial \ac{GSAT} and \ac{ECS} (\cref{fig:03:flato13:a}) and the relationship between \ac{TCR} and \ac{ECS} (\cref{fig:03:flato13:b}) within the \acs{CMIP}5 ensemble. \Cref{fig:03:flato13:a} illustrates that there is no clear connection between \ac{ECS} and the historical or pre-industrial \ac{GSAT}, \ie{} models with high temperatures in the historical or pre-industrial period do no necessarily exhibit a high equilibrium warming in the future or vice versa. On the other hand, \ac{TCR} and \ac{ECS} are in fact well-correlated within the \acs{CMIP}5 ensemble as theoretically expected from \cref{eq:02:tcr_vs_ecs}. However, this figure only shows a linear regression fit between the two variables and not the expected reciprocal relation.

\begin{figure}[t]
  \centering
  \begin{subfigure}[t]{0.34\columnwidth}
    \includegraphics[width=\columnwidth]{ch03_esmvaltool/figs/flato13_942a.pdf}
    \caption{}
    \label{fig:03:flato13:a}
  \end{subfigure}
  ~
  \begin{subfigure}[t]{0.34\columnwidth}
    \includegraphics[width=\columnwidth]{ch03_esmvaltool/figs/flato13_942b.pdf}
    \caption{}
    \label{fig:03:flato13:b}
  \end{subfigure}
  ~
  \begin{subfigure}[t]{0.24\columnwidth}
    \raisebox{7.5mm}{\includegraphics[width=\columnwidth]{
      ch03_esmvaltool/figs/flato13_942_legend.pdf}}
  \end{subfigure}
  \caption{Relationships between climate metrics within the \acs{CMIP}5 climate
    model ensemble. Both panels were created with the \acs{ESMValTool} using
    \emph{recipe\_flato13ipcc.yml} and are similar to \textcite{Flato2013}. (a)
    \Acf{GSAT} versus the \acf{ECS} for \acs{CMIP}5 models for the period
    \range{1961}{1990} (larger symbols) and for pre-industrial control runs
    (smaller symbols). (b) \acf{TCR} versus \acs{ECS} for \acs{CMIP}5 models.
    The black line shows a linear fit between \acs{TCR} and \acs{ECS}.}
  \label{fig:03:flato13}
\end{figure}
